---
title: "Exploratory data analysis of genomic data from breast cancer studies"
author: "Urminder Singh"
date: "December 8, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Our data consists of gene expression data from various samples collected to study breast cancer (Breast Invasive Carcinoma). These samples were collected from a number of different individuals. Then, the samples' RNA were extracted sequenced to get RNA-seq reads from the sample. Using the RNA-seq, the gene expression was estimated. However, we used the data from (Wang et. al. 2018) where the expression data was mapped and normalized again to remove batch effects, which made the dataset comparable across the samples.
Apart from the gene expression data for each sample, we also have associated metadata for each sample. This metadata provides information about the individual, tissue, disease etc.associated with each sample.

Our dataset has expression estimates of around 20,000 genes across 1,092 samples. Out of these 1,092 samples, 982 had tumor and 110 were normal tissues without tumor. We also added the gene metadata to our dataset which describes the gene features and functions. This also includes mutation information of the genes. This information can provide crucial information about how different mutations in the genome can lead to cancer.

Overall, our dataset is GxS matrix where each row corresponds to a gene and each column corresponds to a RNA-seq sample. Additionally, we have metadata for all the rows and all the columns.


# Data

We used data from TCGA. TCGA is a comprehensive repository of human cancer molecular and clinical data, TCGA database has collected clinical and molecular phenotypes of thousands of tumor patients across different tumor types. The TCGA dataset, contains:

* Clinical information about participants
* Metadata about the samples (e.g. the weight of a sample portion, etc.)
* Histopathology slide images from sample portions
* Molecular information derived from the samples (e.g. mRNA/miRNA expression, protein expression, copy number, etc.)



First, we collected data clinical data from TCGA for breast cancer (BRCA) studies. This data contains
Then we downloaded the mutation information for the BRCA studies. This data
Finally, I collected the gene expression data  from (Wang et. al. 2018). This data contains the estimated expression values of genes in the BRCA study. This data was normalized and batch corrected so it is comparable across different TCGA samples.


## Downloading

## Processing


# Analysis

You can also embed plots, for example:

```{r }
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


# Conclusion

# System information

```{r,tidy=TRUE}
sessionInfo()
```
